\begin{examples}[continuous functions]
\lecture{2024-03-07}{Continuity and topology}
    \item All polynomials $p \in \R[x]$ are continuous.
    \item All rational functions $p/q$,
    where $p, q \in \R[x]$ are continuous.
    \item The exponential function $\exp\colon \R \to \R$ is continuous.
    \item The logarithm function $\log\colon (0, \infty) \to \R$
    is continuous.
    \item $x \in \R \mapsto \log_x(a)$, or simply the reciprocal of
    log, is continuous.
\end{examples}
\section{Continuity \& Topology} \label{sec:cont:open-close}
\begin{theorem} \label{thm:cont:open-close}
    Let $(X, d_X)$ and $(Y, d_Y)$ be metric spaces.
    Let $f\colon X \to Y$ be a function.
    Then the following are equivalent:
    \begin{enumerate}
        \item $f$ is continuous on $X$.
        \label{thm:cont:open-close:cont}
        \item $f^{-1}(U)$ is open in $X$ for every open set $U$ in $Y$.
        \label{thm:cont:open-close:open}
        \item $f^{-1}(C)$ is closed in $X$ for every closed set $C$ in $Y$.
        \label{thm:cont:open-close:closed}
    \end{enumerate}
\end{theorem}
\begin{proof}
    For \labelcref{thm:cont:open-close:open}
    $\iff$ \labelcref{thm:cont:open-close:closed},
    use that $f^{-1}(Y \setminus C) = X \setminus f^{-1}(C)$.

    We now show \labelcref{thm:cont:open-close:cont}
    $\iff$ \labelcref{thm:cont:open-close:open},
    Suppose $f$ is continuous on $X$.
    Let $U$ be open in $Y$ and $x \in f^{-1}(U)$.
    Then $f(x) \in U$.
    Since $U$ is open, there exists an $\varepsilon > 0$ such that
    $B_Y(f(x); \varepsilon) \subseteq U$.
    Since $f$ is continuous at $x$, there exists $\delta > 0$ such that
    for all $x' \in X$ with $d_X(x, x') < \delta$,
    $d_Y(f(x), f(x')) < \varepsilon$ so that $f(x') \in U$.
    Then $B_X(x; \delta) \subseteq f^{-1}(U)$.

    Conversely, suppose $f^{-1}(U)$ is open for every open set $U$ in $Y$.
    Let $x \in X$ and $\varepsilon > 0$.
    Choose $U = B_Y(f(x); \varepsilon)$.
    Then $f^{-1}(U)$ is open, so there exists $\delta > 0$ such that
    $B_X(x; \delta) \subseteq f^{-1}(U)$.
    Then for all $x' \in X$ with $d_X(x, x') < \delta$,
    $d_Y(f(x), f(x')) < \varepsilon$.
\end{proof}
\begin{remark}
    Continuous functions need not map open sets to open sets,
    nor closed sets to closed sets.

    $x \mapsto 1$ on $\R$ is continuous, but maps open sets to the closed
    set $\set{1}$.

    $x \mapsto \sin x$ on $\R$ is continuous, but the image of the closed
    set $\N^*$ doesn't contain $0$ ($\pi$ is irrational),
    even though it is a limit point of the image.
\end{remark}

\begin{theorem}[Compactness] \label{thm:cont:open-close:compactness}
    If $f\colon X \to Y$ is continuous and $K \subseteq X$ is compact,
    then $f(K)$ is compact.
\end{theorem}
\begin{proof}
    Let $\set{U_\alpha}$ be an open cover of $f(K)$.
    Each $f^{-1}(U_\alpha)$ is open in $X$ by \cref{thm:cont:open-close}.
    Each $x \in K$ must be in one of these, since otherwise its image
    would not be covered.
    Then $\set{f^{-1}(U_\alpha)}$ is an open cover of $K$.
    Let $\set{f^{-1}(U_{\alpha_i})}$ be a finite subcover of $K$.
    Then $\set{U_{\alpha_i}}$ is a finite subcover of $f(K)$.
\end{proof}
\begin{remarks}
    \item The pre-image of a compact set under a continuous function need
    not be compact.
    \begin{proof}
        Consider $f\colon x \in \R \mapsto \sin x \in \R$.
        Then $f^{-1}(\set{0}) = \Z\pi$ is not compact.
    \end{proof}
    \item Even if the image of every compact set under a function $f$ is
    compact, $f$ need not be continuous.
    \begin{proof}
        Dirichlet's function.
    \end{proof}
\end{remarks}
